Evaluating infectious disease forecasts \\ with allocation scoring rules

### 2024-08-16 alloscore_manuscript/manuscript.tex
knit/latexdiff Evan's changes

% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Tue Aug 13 15:15:48 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
\texttt{CU-select} & 669 & 133 & 774 & 326 \\ 
  \texttt{per-capita} & 865 & - & 1029 & 366 \\ 
  \texttt{COVIDhub-ensemble} & 873 & 159 & 1067 & 438 \\ 
  \texttt{USC-SI\_kJalpha} & 995 & 91 & 1216 & 1097 \\ 
  \texttt{JHUAPL-Gecko} & 1034 & 164 & 1141 & 418 \\ 
  \texttt{MUNI-ARIMA} & 1084 & 169 & 1248 & 440 \\ 
  \texttt{COVIDhub-trained\_ensemble} & 1089 & 169 & 1271 & 823 \\ 
  \texttt{COVIDhub-baseline} & 1175 & 170 & 1317 & 535 \\ 
  \texttt{JHUAPL-Bucky} & 1358 & 102 & 1566 & 1214 \\ 
  \texttt{JHUAPL-SLPHospEns} & 1540 & 129 & 1604 & 1102 \\ 
  \texttt{UVA-Ensemble} & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{For one illustrative week, a comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS), (see Section \ref{sec:ias_examp}). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Tue Aug 13 15:15:52 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
model & MAS & taMWIS & taMWIS rank \\ 
  \hline
\texttt{COVIDhub-ensemble} & 389 & 70 & 2 \\ 
  \texttt{per-capita} & 464 & - & - \\ 
  \texttt{COVIDhub-trained\_ensemble} & 483 & 87 & 4 \\ 
  \texttt{CU-select} & 502 & 81 & 3 \\ 
  \texttt{JHUAPL-SLPHospEns} & 526 & 67 & 1 \\ 
  \texttt{COVIDhub-baseline} & 594 & 93 & 5 \\ 
  \texttt{JHUAPL-Bucky} & 643 & 112 & 7 \\ 
  \texttt{MUNI-ARIMA} & 707 & 116 & 8 \\ 
  \texttt{JHUAPL-Gecko} & 929 & 110 & 6 \\ 
  \texttt{USC-SI\_kJalpha} & 1473 & 155 & 9 \\ 
   \hline
\end{tabular}
\caption{Time averaged mean weighted interval scores (taMWIS) and mean allocation scores (MAS) by model across 13 weeks. These results show the average performance across time for the nine models that submitted forecasts for every week from 2021-11-29 through 2022-02-21, as well
  as for a per-capita allocation. MWIS is not defined for the \texttt{per-capita} allocation which does not use forecasts.} 
\label{tab:multi-week-performance-summary}
\end{table}

@@

AS was computed for a range of $K$ from 200 to 60,000 for forecasts made on December 20, 2020 predicting levels of
hospitalizations on January 3, 2021 as well as for the \texttt{per-capita} allocation (Figure \ref{fig:multi-k}A). These
calculations highlight that AS was highest at at values of $K$ near the observed nationwide total hospital admissions of
19,581 that day. Model ranks by AS were fairly stable across all $K$, and there appears
to be an substantial interval around the observed value over which ranks are constant.

@@

The integrated allocation score (IAS) summarizes allocation scores (AS) across a range of possible values of the
constraint ($K$), allowing one to assign higher weights to more likely values of $K$ (Section
\ref{sec:methods.detailed.integrated_allocation}). 
Calculating exact IAS as defined in equation \eqref{eq:IASsum} is computationally intensive so an approximate IAS, denoted $\hat{S}(F,y)$, was used such that
\begin{align*}
	\hat{S}(F,y) = \sum_{k=1}^{600} S_A(F,y; 200k) p(200k),
\end{align*}
where summation terms were calculated at every 200 beds with an upper bound of 60,000, or four times the estimated available beds.
An approximate IAS was computed for two weight distributions on $K$, one uniform
across the entire range and the other a truncated normal distribution centered at 15,000 (the orange and blue shaded areas in Figure
\ref{fig:multi-k}A, respectively). Both versions of the IAS were correlated with the original AS for $K=15,000$, with
the higher correlation coming from the weighting centered at $K=15,000$ (Figure \ref{fig:multi-k}B). Model rankings
based on the AS and the centered IAS were roughly similar, with the top and bottom three approaches being the same for
both scores (Table \ref{tab:multi-k-scores}).

@@

Probabilistic forecasts of infectious disease outbreaks have typically been evaluated using well-known proper scoring
rules such as the LS, CRPS, and WIS. Often models are ranked according to such a score, but without reference to any
underlying decision problem from which the score might be derivable as a Bayes scoring rule or be otherwise motivated. 
We argue that the value of collaborative forecasting efforts, such as the COVID-19 Forecasting Hub, could be enhanced by
including evaluations of forecasts with
scoring rules that directly consider the success of forecasts in supporting specific decisions. The allocation score was developed for this explicit purpose, evaluating forecasts in the context of the decisions they lead to. The AS is expressed in units that are directly relevant to decisions, which helps not only with ranking forecasts in the context of a specific action, but also allows decision-makers to contextualize the magnitude of differences between forecasts. The benefits of the work presented here generalize beyond public health, to any application of probability forecasts for decision-making, such as economics, where the monetary consequences of using a forecast
can be clearly juxtaposed with standard accuracy measures of the forecast \cite{bannigidadmath2016stock, zhang2019forecasting}, resource allocation in urban planning \citep{burnett2014urban, liang2013resource}, or sustainability initiatives such as water and energy allocation \cite{syme1999defining, gebre2021multi, colett2016using}.

We have demonstrated how tying forecast evaluation to a specific decision problem (e.g., via the allocation
score) can yield model rankings that differ substantively from those based a current standard measure of forecast
accuracy, the WIS, which does not refer to any decision problem that has been clearly connected to outbreak response.
This result aligns with findings in other fields \citep{leitch1991economicForecastEval,
murphy1993whatisagoodforecast, cenesizoglu2012returnPredictionEconValue}. Additionally, we show that an existing
ensemble forecast approach was the only method to outperform (in terms of this new evaluation method) two benchmark
allocation approaches in a hypothetical application, suggesting that there is room for innovation of current
epidemiological modeling techniques that might be spurred on by a reorientation of the field toward practical decision
problems. For example, in order to surpass benchmark allocation scores, forecasters need to devote effort to capturing
the \emph{relative} magnitude of future resource need across different locations.  Success at this task is not, however,
directly rewarded by the WIS (cf. the discussion at the end of Section \ref{sec:methods.related}).

@@

Thirdly, we emphasize that we opted at this stage of
development to explicitly rule out situations where a successful
epidemiological forecast could lead to policy decisions that change the distribution of the predicted outcome $Y$. 
This excludes the fundamentally important context of vaccine allocation for certain pathogens and
spatio-temporal scales, but we note that supply-chain decisions made at a national level in the US determining the state-level
availability of current COVID-19 vaccines are unlikely to affect incidence or exposure levels at a 14-day horizon. In any event,
we are not yet prepared to suggest how valid forecast evaluation using scores given by \eqref{eqn:AS-orc} would proceed when 
both disease burden and unmet need are considered as endogenous variables.

>>>>

% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu Aug 15 23:41:50 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
\texttt{CU-select} & 669 & 133 & 774 & 326 \\ 
  \texttt{per-capita} & 865 & - & 1029 & 366 \\ 
  \texttt{COVIDhub-ensemble} & 873 & 159 & 1067 & 438 \\ 
  \texttt{USC-SI\_kJalpha} & 995 & 91 & 1216 & 1097 \\ 
  \texttt{JHUAPL-Gecko} & 1034 & 164 & 1141 & 418 \\ 
  \texttt{MUNI-ARIMA} & 1084 & 169 & 1248 & 440 \\ 
  \texttt{COVIDhub-trained\_ensemble} & 1089 & 169 & 1271 & 823 \\ 
  \texttt{COVIDhub-baseline} & 1175 & 170 & 1317 & 535 \\ 
  \texttt{JHUAPL-Bucky} & 1358 & 102 & 1566 & 1214 \\ 
  \texttt{JHUAPL-SLPHospEns} & 1540 & 129 & 1604 & 1102 \\ 
  \texttt{UVA-Ensemble} & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{For one illustrative week, a comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS), (see Section \ref{sec:ias_examp}). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu Aug 15 23:41:54 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
model & MAS & taMWIS & taMWIS rank \\ 
  \hline
\texttt{COVIDhub-ensemble} & 389 & 70 & 2 \\ 
  \texttt{per-capita} & 464 & - & - \\ 
  \texttt{COVIDhub-trained\_ensemble} & 483 & 87 & 4 \\ 
  \texttt{CU-select} & 502 & 81 & 3 \\ 
  \texttt{JHUAPL-SLPHospEns} & 526 & 67 & 1 \\ 
  \texttt{COVIDhub-baseline} & 594 & 93 & 5 \\ 
  \texttt{JHUAPL-Bucky} & 643 & 112 & 7 \\ 
  \texttt{MUNI-ARIMA} & 707 & 116 & 8 \\ 
  \texttt{JHUAPL-Gecko} & 929 & 110 & 6 \\ 
  \texttt{USC-SI\_kJalpha} & 1473 & 155 & 9 \\ 
   \hline
\end{tabular}
\caption{Time averaged mean weighted interval scores (taMWIS) and mean allocation scores (MAS) by model across 13 weeks. These results show the average performance across time for the nine models that submitted forecasts for every week from 2021-11-29 through 2022-02-21, as well
  as for a per-capita allocation. MWIS is not defined for the \texttt{per-capita} allocation which does not use forecasts.} 
\label{tab:multi-week-performance-summary}
\end{table}

@@

AS was computed for a range of $K$ from 200 to 60,000 at increments of 200 for forecasts made on December 20, 2020 predicting levels of
hospitalizations on January 3, 2021 as well as for the \texttt{per-capita} allocation (Figure \ref{fig:multi-k}A). These
calculations highlight that AS was highest at at values of $K$ near the observed nationwide total hospital admissions of
19,581 that day. Model ranks by AS were fairly stable across all $K$, and there appears
to be an substantial interval around the observed value over which ranks are constant.

@@

The integrated allocation score (IAS) summarizes allocation scores (AS) across a range of possible values of the
constraint ($K$), allowing one to assign higher weights to more likely values of $K$ (Section
\ref{sec:methods.detailed.integrated_allocation}). 
IAS was computed for two weight distributions on the same grid of values of $K$ used in the sensitivity analysis above,
one uniform across the entire range and the other with weights proportional to a truncated normal distribution centered at 15,000
(the orange and blue shaded areas in Figure
\ref{fig:multi-k}A, respectively).
With these weight distributions, the IAS defined in Equation~\eqref{eqn:IAS_definition} can be computed as the sum
$$S_{IAS}(F, y) = \sum_K S_A(F,y; K) p(K)$$

Both versions of the IAS were correlated with the original AS for $K=15,000$, with
the higher correlation coming from the weighting centered at $K=15,000$ (Figure \ref{fig:multi-k}B). Model rankings
based on the AS and the centered IAS were roughly similar, with the top and bottom three approaches being the same for
both scores (Table \ref{tab:multi-k-scores}).

@@

Probabilistic forecasts of infectious disease outbreaks have typically been evaluated using well-known proper scoring
rules such as the LS, CRPS, and WIS. Often models are ranked according to such a score, but without reference to any
underlying decision problem from which the score might be derivable as a Bayes scoring rule or be otherwise motivated. 
We argue that the value of collaborative forecasting efforts, such as the COVID-19 Forecasting Hub, could be enhanced by
including evaluations of forecasts with
scoring rules that directly consider the success of forecasts in supporting specific decisions. The allocation score was developed for this purpose. The AS is expressed in units that are directly relevant to decisions, which helps not only with ranking forecasts in the context of a specific action, but also allows decision-makers to contextualize the magnitude of differences between forecast scores. The benefits of the approach to forecast evaluation discussed here generalize beyond public health, to any application of probability forecasts for decision-making, such as economics, where the monetary consequences of using a forecast
can be clearly juxtaposed with standard accuracy measures of the forecast \citep{bannigidadmath2016stock, zhang2019forecasting}, resource allocation in urban planning \citep{burnett2014urban, liang2013resource}, or sustainability initiatives such as water and energy allocation \citep{syme1999defining, gebre2021multi, colett2016using}.

We have demonstrated how tying forecast evaluation to a specific decision problem (e.g., via the allocation
score) can yield model rankings that differ substantively from those based a current standard measure of forecast
accuracy, the WIS, which does not refer to any decision problem that has been clearly connected to outbreak response.
This result aligns with findings in other fields \citep{leitch1991economicForecastEval,
murphy1993whatisagoodforecast, cenesizoglu2012returnPredictionEconValue}. Additionally, we show that an existing
ensemble forecast approach was the only method to outperform (in terms of this new evaluation method) two benchmark
allocation approaches in a hypothetical application, suggesting that there is room for innovation of current
epidemiological modeling techniques that might be spurred on by a reorientation of the field toward practical decision
problems. For example, in order to surpass benchmark allocation scores, forecasters need to devote effort to capturing
the \emph{relative} magnitude of future resource need across different locations.  Success at this task is not, however,
directly rewarded by the WIS (see the discussion at the end of Section \ref{sec:methods.related}).

@@

Thirdly, we emphasize that we opted at this stage of
development to explicitly rule out situations where a successful
epidemiological forecast could lead to policy decisions that change the distribution of the predicted outcome $Y$. 
For example, this excludes the important problem of allocation of doses of a preventative vaccine.
We have not developed a formal approach to forecast evaluation using scores given by \eqref{eqn:AS-orc} when both
disease burden and unmet need are considered as endogenous variables. Such an approach would require some form of
accounting for the causal effect of the allocation decision on observed measures of disease burden, e.g. using the available observed data to estimate the (counterfactual)
number of cases or deaths averted from one or another allocation strategy.
